---
title: "Economic and Public Health Impact of Severe Weather"
author: "Katelyn Billings"
date: "January 14, 2015"
output: 
    html_document:
        keep_md: TRUE
---

###Synopsis

This report aims to identify the types of storms that cause the most harm to 
public health and have the greatest economic consequences. To investigate this 
matter, storm data from the U.S. National Oceanic and Atmospheric Administration
were analyzed. The data spanned the years 1950-2011 and documented when and 
where the storms occurred and estimated the number of fatalities and injuries 
and any crop or property damages. From the data it was identified that 
**tornados are the greatest cause of harm to public health** (as defined as the 
number of fatalities and injuries per year) while **hurricanes and hurricane/
typhoons have the greatest economic consequences** (as defined by the cost of 
property and crop damage per year).

###Key Questions

The following analysis addresses the following questions:

1. Across the United States, which types of events are most harmful with respect 
to population health?

2. Across the United States, which types of events have the greatest economic 
consequences?

This report should help someone prioritize resources for different types of 
events.

###Data Processing

From the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm 
database, [data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) 
on major storms and weather events in the United States were obtained for the 
period spanning 1950-2011. The data documents when and where the storms occur,
as well as estimates of fatalities, injuries and any property or crop damage.

#### Data Download

To begin the analysis, it is important to check if the 
[storm data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) 
is already in the working directory. If it is not in the working directory, the 
file is downloaded and unzipped.

```{r Get Data}
if(!file.exists("./repdata-data-StormData.csv")){
    fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(fileUrl, destfile = "./downloadedData", method = "curl")
    dateDownloaded <<- date()
    unzip("./downloadedData")
}
```


####Read in the Data

The raw comma-separated-value file included in the zip archive must then be read
into memory. NA values are indicated by the string "NA" or empty strings.

```{r Read Data, cache=TRUE}
originalStormData <- read.csv("./repdata-data-StormData.csv", 
                              na.strings = c("NA", ""))
```

A quick look at the data set:

```{r Structure Original Data}
str(originalStormData)
```

####Clean the data

Any dependencies that are necessary for data analysis must be loaded.

```{r Load Dependencies}
library(lubridate, warn.conflicts = FALSE, quietly = TRUE)
library(plyr, warn.conflicts = FALSE, quietly = TRUE)
library(dplyr, warn.conflicts = FALSE, quietly = TRUE)
```

Since this analysis is only concerned about which types of storms have the 
greatest economic consequences and impact on public health, only a subset of the 
data is retained: the storm type, fatalities/injuries, and property/crop 
damages.

```{r Subset data}

stormData <- select(originalStormData, date = BGN_DATE, storm.type = EVTYPE, 
                    fatalities = FATALITIES, injuries = INJURIES, 
                    property.damage = PROPDMG, property.exp = PROPDMGEXP, 
                    crop.damage = CROPDMG, crop.exp = CROPDMGEXP)
stormData$date <- mdy_hms(stormData$date)
num.years <- length(unique(year(stormData$date)))

```

The structure of the data is now:

```{r Data Structure}
str(stormData)
```

Property and crop damage have a corresponding exponent/multiplier associated 
with them that must be accounted for. The levels and frequencies of the property 
and crop exponents are:

```{r Summary of Exponent Factors}
table(stormData$property.exp, useNA = "ifany")
table(stormData$crop.exp, useNA = "ifany")
```

It is unclear what factor levels -, ?, + mean. Moreover, this subset of data
represents a very small proportion of the overall data. As a result, the data 
that contain these factor levels will be removed:

```{r Remove Unclear Factor Levels}
stormData <- stormData[grep("\\-|\\?|\\+", stormData$property.exp, 
                       invert = TRUE), ]
stormData <- stormData[grep("\\?", stormData$crop.exp, invert = TRUE), ]
```

The character based factor levels need to be reassigned to the corresponding 
numeric counterparts (eg. h/H stands for hundred so it gets reassigned to 2 
since 100 = 10^2; likewise, k/K = 3, m/M = 6, b/B = 9, NA = 0)

```{r Reassign factor levels}
# Reassign levels for property.exp, convert to a numeric
stormData$property.exp <- mapvalues(stormData$property.exp, 
                                    from = c("h", "H", "K", "m", "M", "B"),
                                    to   = c("2", "2", "3", "6", "6", "9"))
stormData[which(is.na(stormData$property.exp)), 6] <- "0"
stormData$property.exp <- as.numeric(as.character(stormData$property.exp))

# Reassign levels for crop.exp, convert to a numeric
stormData$crop.exp <- mapvalues(stormData$crop.exp, 
                                from = c("k", "K", "m", "M", "B"),
                                to   = c("3", "3", "6", "6", "9"))
stormData[which(is.na(stormData$crop.exp)), 8] <- "0"
stormData$crop.exp <- as.numeric(as.character(stormData$crop.exp))
```

A verification that the property and crop exponents were successfully 
reassigned:

```{r Summary Reassigned Exponents}
table(stormData$property.exp, useNA = "ifany")
table(stormData$crop.exp, useNA = "ifany")
```

The property and crop damages must then be scaled by the corresponding 
exponents. The .exp variables are then dropped.

```{r Scale Damages}
stormData <- mutate(stormData, 
                    property.damage = property.damage * 10^property.exp,
                    crop.damage = crop.damage * 10^property.exp) %>%
             select(-c(property.exp, crop.exp))
```

The data now looks like:

```{r Revised Data Summary}
str(stormData)
summary(stormData)
```

###Results

####Impact on Public Health

The first goal of this analysis is to identify which types of events are most 
harmful with respect to population health. Since the data spans `r num.years` 
years and this analysis looks to inform decisions on a yearly basis, all storm 
data is scaled to a per year amount (total amount/number ofyears).

A look at what types of events occur most often on a per year basis:

```{r Most Frequent Events}
num.events <- as.data.frame(table(as.character(stormData$storm.type)))
num.events$Freq <- round(num.events$Freq/num.years) 
arrange(num.events, desc(Freq))[1:10, ]
```

The top ten causes of fatalities/year are: 

```{r fatalities}
# Get top ten causes of fatalities/year
health.fatalities <- group_by(stormData, storm.type) %>%
                     summarize(fatalities = sum(fatalities, na.rm = TRUE))
health.fatalities$fatalities <- round(health.fatalities$fatalities/num.years)
topten.fatalities <- arrange(health.fatalities, desc(fatalities))[1:10, ]
topten.fatalities
```

The top ten causes of injuries/year are:

```{r injuries}
# Get top ten causes of injuries/year
health.injuries <- group_by(stormData, storm.type) %>%
                   summarize(injuries = sum(injuries, na.rm = TRUE))
health.injuries$injuries <- round(health.injuries$injuries/num.years)
topten.injuries <- arrange(health.injuries, desc(injuries))[1:10, ]
topten.injuries
```

The top ten causes of all health problems per year (fatalities + injuries) are:

```{r total problems}
# Get top ten causes of health problems
health.problems <- group_by(stormData, storm.type) %>%
                   summarize(problems = sum(injuries + fatalities, na.rm = TRUE))
health.problems$problems <- round(health.problems$problems/num.years)
topten.problems <- arrange(health.problems, desc(problems))[1:10, ]
topten.problems
```

```{r fig.width = 12, fig.height = 6}
# Display information in graphical form
library(RColorBrewer)
par(mfcol=c(1,1), mar = c(7,6,4,4), las = 2, cex.main = 1, cex.axis = 0.9,
    mgp = c(5, 1, 0))
with(topten.problems, barplot(problems, names.arg = storm.type,
                              xlab = "Storm Type",
                              ylab = "Number of Injuries + Fatalities Per Year",
                              main = paste0("Top 10 Causes of Injuries and ",
                                            "Fatalities Combined"),
                              cex.names = 0.6,
                              col = brewer.pal(10, "Set3")))
```

The plot above indicates that `r topten.problems[1,1]` is the most harmful
type of event to personal health at `r topten.problems[1,2]` injuries and 
fatalities per year. The level of harm drops off afterwards, with
`r topten.problems[2,1]` being the second leading cause of harm with 
`r topten.problems[2,2]` health incidents.


####Greatest Economic Consequences

The second goal of this analysis is to identify the types of storms that have
the greatest economic consequences. For this analysis, economic consequences
refers to property and crop damage (in millions of dollars). Once again all
storm data is scaled to a per year basis. 

The top ten causes of property damage per year ($, millions) are:

```{r property damage}
# Get top ten causes of property damage ($, billions)
econ.property <- group_by(stormData, storm.type) %>%
                 summarize(property.damage = sum(property.damage/10^6, 
                                                 na.rm = TRUE))
econ.property$property.damage <- round(econ.property$property.damage/num.years)
topten.property <- arrange(econ.property, desc(property.damage))[1:10, ]
topten.property
```

The top ten causes of crop damage per year ($, millions) are:

```{r crop damage}
# Get top ten causes of crop damage ($, millions)
econ.crop <- group_by(stormData, storm.type) %>%
             summarize(crop.damage = sum(crop.damage/10^6, na.rm = TRUE))
econ.crop$crop.damage <- round(econ.crop$crop.damage/num.years)
topten.crop <- arrange(econ.crop, desc(crop.damage))[1:10, ]
topten.crop
```

The top ten causes of property and crop damage combined per year ($, millions) 
are:

```{r total damage}
# Get the top ten causes of total damage ($, millions)
econ.total <- group_by(stormData, storm.type) %>%
              summarize(total.damage = sum(property.damage + crop.damage, 
                                           na.rm = TRUE)/10^6)
econ.total$total.damage <- round(econ.total$total.damage/num.years)
topten.total <- arrange(econ.total, desc(total.damage))[1:10, ]
topten.total
```

```{r fig.width = 12, fig.height = 6}
# Display causes of property/crop damage in graphical form
par(mfcol=c(1,1), mar = c(7,6,4,4), las = 2, cex.main = 1, cex.axis = 0.9, 
    mgp = c(5, 1, 0))
with(topten.total, barplot(total.damage, names.arg = storm.type,
                           xlab = "Storm Type",
                           ylab = paste0("Property + Crop Damage Per Year ",
                                         "\n($, millions)"),
                           main = "Top 10 Causes of Property and Crop Damage",
                           cex.names = 0.6,
                           col = brewer.pal(10, "Set3")))
```

The plot above indicates that `r topten.total[1,1]` causes the most damage 
economically with `r topten.total[1,2]` million dollars of damage to property
and crops per year. `r topten.total[2,1]` is the second most costly type of
storm with `r topten.total[2,2]` million dollars of damage to property and
crops per year.