---
title: "Economic and Public Health Impact of Severe Weather"
author: "Katelyn Billings"
date: "January 14, 2015"
output: 
    html_document:
        keep_md: TRUE
---

###Synopsis

###Key Questions

The data analysis addresses the following questions:

1. Across the United States, which types of events (as indicated in the EVTYPE 
variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic 
consequences?

The analysis should help someone prioritize resources for different types of 
events.

###Data Processing

From the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm 
database, we obtained data on major storms and weather events in the United 
States, including when and where they occur, as well as estimates of any 
fatalities, injuries, and property damage.

#### Data Download

We first check if the data has already been downloaded from the 
[NOAA Storm Database](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).
If it does not exist, the file is downloaded and unzipped.

```{r Get Data}
if(!file.exists("./repdata-data-StormData.csv")){
    fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(fileUrl, destfile = "./downloadedData", method = "curl")
    dateDownloaded <<- date()
    unzip("./downloadedData")
}
```


####Read in the Data

We then read in the raw comma-separated-value file included in the zip archive.
NA values are indicated by the string "NA" or empty strings.

```{r Read Data, cache=TRUE}
originalStormData <- read.csv("./repdata-data-StormData.csv", 
                              na.strings = c("NA", ""))
```

A quick look at the data set:

```{r}
str(originalStormData)
```

####Clean the data

We need to load any dependencies that will be required for data analysis.

```{r}
library(lubridate, warn.conflicts = FALSE, quietly = TRUE)
library(plyr, warn.conflicts = FALSE, quietly = TRUE)
library(dplyr, warn.conflicts = FALSE, quietly = TRUE)
```

Since we are only concerned about which types of storms have the greatest 
economic consequences and impact on public health, only a subset of the data 
will be retained: the storm type, fatalities/injuries, and property/crop damages.

```{r Working Data}

stormData <- select(originalStormData, date = BGN_DATE, storm.type = EVTYPE, 
                    fatalities = FATALITIES, injuries = INJURIES, 
                    property.damage = PROPDMG, property.exp = PROPDMGEXP, 
                    crop.damage = CROPDMG, crop.exp = CROPDMGEXP)
stormData$date <- mdy_hms(stormData$date)

```

The structure of the data is now:

```{r}
str(stormData)
```

We now need to take into account the fact that property and crop damage have a 
corresponding exponent/multiplier associated with them. Let's look at the levels
and frequencies of the property and crop exponents:

```{r}
table(stormData$property.exp, useNA = "ifany")
table(stormData$crop.exp, useNA = "ifany")
```

It is unclear what factor levels -, ?, + mean so we will remove the data that
contain these factor levels:

```{r}
stormData <- stormData[grep("\\-|\\?|\\+", stormData$property.exp, 
                       invert = TRUE), ]
stormData <- stormData[grep("\\?", stormData$crop.exp, invert = TRUE), ]
```

We now need to reassign the character levels to their corresponding numeric 
counterparts: h/H = 2, k/K = 3, m/M = 6, b/B = 9, NA = 0

```{r}
# Reassign levels for property.exp, convert to a numeric
stormData$property.exp <- mapvalues(stormData$property.exp, 
                                    from = c("h", "H", "K", "m", "M", "B"),
                                    to   = c("2", "2", "3", "6", "6", "9"))
stormData[which(is.na(stormData$property.exp)), 6] <- "0"
stormData$property.exp <- as.numeric(as.character(stormData$property.exp))

# Reassign levels for crop.exp, convert to a numeric
stormData$crop.exp <- mapvalues(stormData$crop.exp, 
                                from = c("k", "K", "m", "M", "B"),
                                to   = c("3", "3", "6", "6", "9"))
stormData[which(is.na(stormData$crop.exp)), 8] <- "0"
stormData$crop.exp <- as.numeric(as.character(stormData$crop.exp))
```

Let's make sure the property and crop exponents were successfully reassigned:

```{r}
table(stormData$property.exp, useNA = "ifany")
table(stormData$crop.exp, useNA = "ifany")
```

Finally, we need to multiply the property damage by their corresponding 
exponents then remove the .exp variables.

```{r}
stormData <- mutate(stormData, 
                    property.damage = property.damage * 10^property.exp,
                    crop.damage = crop.damage * 10^property.exp) %>%
             select(-c(property.exp, crop.exp))
```

Our data now looks like:

```{r}
str(stormData)
summary(stormData)
```

###Results

####Impact on Public Health

We need to figure out which types of events are most harmful with respect to 
population health. First let's look at what types of events occur most often.

```{r}
num.events <- as.data.frame(table(as.character(stormData$storm.type)))
arrange(num.events, desc(Freq))[1:10, ]
```

Let's look at the top ten causes of fatalities and injuries: 

```{r fatalities}
# Get top ten causes of fatalities
health.fatalities <- group_by(stormData, storm.type) %>%
                     summarize(fatalities = sum(fatalities, na.rm = TRUE))
topten.fatalities <- arrange(health.fatalities, desc(fatalities))[1:10, ]
topten.fatalities
```

```{r injuries}
# Get top ten causes of injuries
health.injuries <- group_by(stormData, storm.type) %>%
                   summarize(injuries = sum(injuries, na.rm = TRUE))
topten.injuries <- arrange(health.injuries, desc(injuries))[1:10, ]
topten.injuries
```

Finally, let's look at the top ten causes of all health problems (fatalities +
injuries) and display the results in a bar plot:

```{r total problems}
# Get top ten causes of health problems
health.problems <- group_by(stormData, storm.type) %>%
                   summarize(problems = sum(injuries + fatalities, na.rm = TRUE))
topten.problems <- arrange(health.problems, desc(problems))[1:10, ]
topten.problems
```

```{r fig.width = 12, fig.height = 6}
# Display information in graphical form
library(RColorBrewer)
par(mfcol=c(1,1), mar = c(7,6,4,4), las = 2, cex.main = 1, cex.axis = 0.9,
    mgp = c(5, 1, 0))
with(topten.problems, barplot(problems, names.arg = storm.type,
                              xlab = "Storm Type",
                              ylab = "Number of Injuries + Fatalities",
                              main = "Top 10 Causes of Injuries and Fatalities Combined",
                              cex.names = 0.6,
                              col = brewer.pal(10, "Set3")))
```

From the plot above, we see that `r topten.problems[1,1]` is the most harmful
type of event to personal health at `r topten.problems[1,2]` injuries and 
fatalities combined. The level of harm drops off afterwards, with
`r topten.problems[2,1]` being the second leading cause of harm with 
`r topten.problems[2,2]` health incidents.


####Greatest Economic Consequences

In terms of economic consequences, we are concerned with the amount of property
and crop damage caused by a given storm type. Let's first look at the top ten 
causes of property and crop damage separately ($, billions):

```{r property damage}
# Get top ten causes of property damage ($, billions)
econ.property <- group_by(stormData, storm.type) %>%
                 summarize(property.damage = sum(property.damage/10^9, 
                                                 na.rm = TRUE))
topten.property <- arrange(econ.property, desc(property.damage))[1:10, ]
topten.property
```

```{r crop damage}
# Get top ten causes of crop damage ($, billions)
econ.crop <- group_by(stormData, storm.type) %>%
             summarize(crop.damage = sum(crop.damage/10^9, na.rm = TRUE))
topten.crop <- arrange(econ.crop, desc(crop.damage))[1:10, ]
topten.crop
```

Finally let's look at the top ten causes of property and crop damage combined
($, billions) and display the results in a bar plot:

```{r total damage}
# Get the top ten causes of total damage ($, billions)
econ.total <- group_by(stormData, storm.type) %>%
              summarize(total.damage = sum(property.damage + crop.damage, 
                                           na.rm = TRUE)/10^9)
topten.total <- arrange(econ.total, desc(total.damage))[1:10, ]
topten.total
```

```{r fig.width = 12, fig.height = 6}
# Display causes of property/crop damage in graphical form
par(mfcol=c(1,1), mar = c(7,6,4,4), las = 2, cex.main = 1, cex.axis = 0.9, 
    mgp = c(5, 1, 0))
with(topten.total, barplot(total.damage, names.arg = storm.type,
                           xlab = "Storm Type",
                           ylab = "Property + Crop Damage ($, billions)",
                           main = "Top 10 Causes of Property and Crop Damage",
                           cex.names = 0.6,
                           col = brewer.pal(10, "Set3")))
```

From the plot above, we see that `r topten.total[1,1]` causes the most damage 
economically with `r topten.total[1,2]` billion dollars of damage to property
and crops. `r topten.total[2,1]` is the second most costly type of
storm with `r topten.total[2,2]` billion dollars of damage to property and
crops.