---
title: "Economic and Public Health Impact of Severe Weather"
author: "Katelyn Billings"
date: "January 2015"
output: 
    html_document:
        keep_md: TRUE
---

###Synopsis

This report aims to identify the types of storms that cause the most harm to 
public health and have the greatest economic consequences. To investigate this 
matter, storm data from the U.S. National Oceanic and Atmospheric Administration
were analyzed. The data documents when and where storms occurred and estimates 
the number of fatalities and injuries, as well as any crop or property damages. 
Only the data from 1996-2011 were used for analysis due to incomplete record
keeping in years prior to 1996. From the data it was identified that 
**tornados are the greatest cause of harm to public health** (as defined as the 
number of fatalities and injuries per year) while **hurricanes and hurricane/
typhoons have the greatest economic consequences** (as defined by the cost of 
property and crop damage per year).

###Key Questions

The following analysis addresses the following questions:

1. Across the United States, which types of events are most harmful with respect 
to population health?

2. Across the United States, which types of events have the greatest economic 
consequences?

This report should help someone prioritize resources for different types of 
events.

###Data Processing

From the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm 
database, [data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) 
on major storms and weather events in the United States were obtained for the 
period spanning 1950-2011. Only data from 1996-2011 will be used for data 
analysis however, due to incomplete record keeping prior to 1996. The data 
documents when and where the storms occur, as well as estimates of fatalities, 
injuries and any property or crop damage. More information on the database can 
be found here:

-National Weather Service Storm Database [Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

-National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)


#### Data Download

To begin the analysis, it is important to check if the 
[storm data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) 
is already in the working directory. If it is not in the working directory, the 
file is downloaded and unzipped.

```{r Get Data}
if(!file.exists("./repdata-data-StormData.csv")){
    fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(fileUrl, destfile = "./downloadedData", method = "curl")
    dateDownloaded <<- date()
    unzip("./downloadedData")
}
```


####Read in the Data

The raw comma-separated-value file included in the zip archive must then be read
into memory. NA values are indicated by the string "NA" or empty strings.

```{r Read Data, cache=TRUE}
originalStormData <- read.csv("./repdata-data-StormData.csv", 
                              na.strings = c("NA", ""))
```

A quick look at the data set:

```{r Structure Original Data}
str(originalStormData)
```

####Clean the data

Any dependencies that are necessary for data analysis must be loaded.

```{r Load Dependencies}
library(lubridate, warn.conflicts = FALSE, quietly = TRUE)
library(plyr, warn.conflicts = FALSE, quietly = TRUE)
library(dplyr, warn.conflicts = FALSE, quietly = TRUE)
```

Since this analysis is only concerned about which types of storms have the 
greatest economic consequences and impact on public health, only a subset of the 
data is retained: the storm type, fatalities/injuries, property/crop damages, 
and the corresponding scaling factor for the damages. 

Additionally, only the data reported from 1996-2011 will be kept due to 
incomplete record keeping prior to these years. 

```{r Subset data}

stormData <- select(originalStormData, date = BGN_DATE, storm.type = EVTYPE, 
                    fatalities = FATALITIES, injuries = INJURIES, 
                    property.damage = PROPDMG, property.exp = PROPDMGEXP, 
                    crop.damage = CROPDMG, crop.exp = CROPDMGEXP)
stormData$date <- mdy_hms(stormData$date)
stormData <- filter(stormData, year(date) >= 1996)
num.years <- length(unique(year(stormData$date)))

```

The structure of the data is now:

```{r Data Structure}
str(stormData)
```

Property and crop damage have a corresponding exponent/multiplier associated 
with them that must be accounted for. The levels and frequencies of the property 
and crop exponents are:

```{r Summary of Exponent Factors}
table(stormData$property.exp, useNA = "ifany")
table(stormData$crop.exp, useNA = "ifany")
```

The character based factor levels need to be reassigned to the corresponding 
numeric counterparts (eg. K stands for thousand so it gets reassigned to 3 
since 1000 = 10^3; likewise, M = 6, B = 9, NA = 0)

```{r Reassign factor levels}
# Reassign levels for property.exp, convert to a numeric
stormData$property.exp <- mapvalues(stormData$property.exp, 
                                    from = c("K", "M", "B"),
                                    to   = c("3", "6", "9"))
stormData[which(is.na(stormData$property.exp)), 6] <- "0"
stormData$property.exp <- as.numeric(as.character(stormData$property.exp))

# Reassign levels for crop.exp, convert to a numeric
stormData$crop.exp <- mapvalues(stormData$crop.exp, 
                                from = c("K", "M", "B"),
                                to   = c("3", "6", "9"))
stormData[which(is.na(stormData$crop.exp)), 8] <- "0"
stormData$crop.exp <- as.numeric(as.character(stormData$crop.exp))
```

A verification that the property and crop exponents were successfully 
reassigned:

```{r Summary Reassigned Exponents}
table(stormData$property.exp, useNA = "ifany")
table(stormData$crop.exp, useNA = "ifany")
```

The property and crop damages must then be scaled by the corresponding 
exponents. The .exp variables are then dropped.

```{r Scale Damages}
stormData <- mutate(stormData, 
                    property.damage = property.damage * 10^property.exp,
                    crop.damage = crop.damage * 10^property.exp) %>%
             select(-c(date, property.exp, crop.exp))
```

The data can then be combined by storm type, averaging the results across the 
subset of years originally selected:

```{r}
summedData <- ddply(stormData, .(storm.type), summarize,
                    injuries.yr = sum(injuries)/num.years,
                    fatalities.yr = sum(fatalities)/num.years,
                    totalhealth.yr = sum(injuries + fatalities)/num.years,
                    propertydmg.yr = sum(property.damage)/num.years,
                    cropdmg.yr = sum(crop.damage)/num.years,
                    totaldmg.yr = sum(property.damage + crop.damage)/num.years)
```

The data now looks like:

```{r Revised Data Summary}
str(summedData)
summary(summedData)
```

###Results

####Impact on Public Health

The first goal of this analysis is to identify which types of events are most 
harmful with respect to population health. This analysis identifies the leading
causes of injuries and fatalities both separately and combined on a per year 
basis. 

A look at what types of events occur most often on a per year basis:

```{r Most Frequent Events}
num.events <- as.data.frame(table(as.character(stormData$storm.type)))
summedData <- merge(summedData, num.events, by.x = "storm.type", by.y = "Var1")
summedData <- mutate(summedData, freq.yr = Freq/num.years) %>%
              select(-Freq)
by.events <- select(summedData, storm.type, freq.yr)
topten.events <- arrange(by.events, desc(freq.yr))[1:10, ]
topten.events[1:10, ]
```

The top ten causes of fatalities/year are: 

```{r fatalities}
# Get top ten causes of fatalities/year
by.fatalities <- select(summedData, storm.type, fatalities.yr, freq.yr)
topten.fatalities <- arrange(by.fatalities, desc(fatalities.yr))[1:10, ]
topten.fatalities <- mutate(topten.fatalities, 
                            fatalities.event = round(fatalities.yr/freq.yr, 4))
topten.fatalities[ ,c(1,2,4)]
```

The top ten causes of injuries/year are:

```{r injuries}
# Get top ten causes of injuries/year
by.injuries <- select(summedData, storm.type, injuries.yr, freq.yr)
topten.injuries <- arrange(by.injuries, desc(injuries.yr))[1:10, ]
topten.injuries <- mutate(topten.injuries,
                          injuries.event = round(injuries.yr/freq.yr, 4))
topten.injuries[ , c(1,2,4)]
```

The top ten causes of all health problems (fatalities + injuries) per year are:

```{r total problems}
# Get top ten causes of health problems
by.health <- select(summedData, storm.type, totalhealth.yr, freq.yr)
topten.health <- arrange(by.health, desc(totalhealth.yr))[1:10, ]
topten.health <- mutate(topten.health, 
                        totalhealth.event = round(totalhealth.yr/freq.yr, 4))
topten.health[ , c(1,2,4)]
```

A graphical look at the top ten events leading to the most fatalities and 
injuries:

```{r fig.width = 12, fig.height = 6, fig.align = 'center'}
# Display information in graphical form
library(RColorBrewer)
par(mfrow = c(1, 2), las = 1, mar = c(5, 4, 6, 2))
with(topten.health, 
     barplot(totalhealth.yr, 
             xlab = "Storm Type",
             ylab = "Fatalities + Injuries Per Year",
             main = paste0("Cumulative Fatalities and Injuries \nper Year"),
             font.main = 1,
             cex.names = 0.6,
             col = brewer.pal(10, "Set3"),
             legend.text = storm.type))     
with(topten.health,     
     plot(freq.yr, totalhealth.event,
          main = paste0("Events per Year vs \nHealth Problems per Event"),
          font.main = 1,
          xlab = "Events Per Year",
          ylab = "Fatalities and Injuries Per Event",
          pch = 19,
          col = brewer.pal(10, "Set3")))
with(topten.health, legend("topright", legend = storm.type, pch = 19, 
                           col = brewer.pal(10, "Set3"), cex = 1))
mtext("Top 10 Events Leading to the Most Fatalities + Injuries", side = 3, 
      line = -1, outer = TRUE, cex = 1.2, font = 2)
```

The resulting graph is quite interesting: of the top ten events that cause the 
most harm to public health, `r topten.health[1,1]` causes the most number of 
health problems per year with `r topten.health[1,2]` health problems, yet only 
causes `r topten.health[1,3]` health problems per event in the year. 

In fact, for the majority of the events that cause the most harm to public 
health each year, the corresponding number fatalities + injuries per events is 
close to zero. This suggests that the individual events cause minimal harm, but 
because there are so many events each year, the sum of all harm becomes 
substantial.

####Greatest Economic Consequences

The second goal of this analysis is to identify the types of storms that have
the greatest economic consequences. For this analysis, economic consequences
refers to property and crop damage (in millions of dollars).

The top ten causes of property damage per year ($, millions) are:

```{r property damage}
# Get top ten causes of property damage ($, millions)
by.property <- select(summedData, storm.type, propertydmg.yr, freq.yr)
topten.property <- arrange(by.property, desc(propertydmg.yr))[1:10, ]
topten.property <- mutate(topten.property, 
                          propertydmg.yr = propertydmg.yr/10^6,
                          propertydmg.event = round(propertydmg.yr/freq.yr, 4))
topten.property[ , c(1,2,4)]
```

The top ten causes of crop damage per year ($, millions) are:

```{r crop damage}
# Get top ten causes of crop damage ($, millions)
by.crop <- select(summedData, storm.type, cropdmg.yr, freq.yr)
topten.crop <- arrange(by.crop, desc(cropdmg.yr))[1:10, ]
topten.crop <- mutate(topten.crop,
                      cropdmg.yr = cropdmg.yr/10^6,
                      cropdmg.event = round(cropdmg.yr/freq.yr, 4))
topten.crop[ , c(1,2,4)]
```

The top ten causes of property and crop damage combined per year ($, millions) 
are:

```{r total damage}
# Get the top ten causes of total damage ($, millions)
by.totaldmg <- select(summedData, storm.type, totaldmg.yr, freq.yr)
topten.totaldmg <- arrange(by.totaldmg, desc(totaldmg.yr))[1:10, ]
topten.totaldmg <- mutate(topten.totaldmg,
                          totaldmg.yr = totaldmg.yr/10^6,
                          totaldmg.event = round(totaldmg.yr/freq.yr, 4))
topten.totaldmg[ , c(1,2,4)]
```

The top ten causes of property and crop damage per event ($, millions)

```{r fig.width = 12, fig.height = 6, fig.align = "center"}
# Display causes of property/crop damage in graphical form
par(mfrow = c(1, 2), las = 1, mar = c(5, 7, 6, 2), mgp = c(4, 1, 0))
with(topten.totaldmg, 
     barplot(totaldmg.yr, 
             xlab = "Storm Type",
             ylab = "Damages Per Year ($, millions)",
             main = paste0("Cumulative Damages per Year"),
             font.main = 1,
             cex.names = 0.6,
             col = brewer.pal(10, "Set3"),
             legend.text = storm.type))
with(topten.totaldmg, plot(freq.yr, totaldmg.event,
                           main = "Events per Year vs Damages per Event",
                           font.main = 1,
                           xlab = "Events per Year",
                           xlim = c(0, max(freq.yr)*1.2),
                           ylab = "Damages Per Event ($, millions)",
                           ylim = c(0, max(totaldmg.event)*1.2),
                           pch = 19,
                           col = brewer.pal(10, "Set3")))
with(topten.totaldmg, legend("topright", legend = storm.type, pch = 19,
                             col = brewer.pal(10, "Set3"), cex = 1))
mtext("Top 10 Events Leading to the Most Property and Crop Damages", side = 3,
      line = -1, outer = TRUE, cex = 1.2, font = 2)
```

This graph indicates that `r topten.totaldmg[1,1]` has the greatest economic 
consequences with `r topten.totaldmg[1,2]` million dollars of damage each year
and `r topten.totaldmg[1,4]` million dollars of damage each event to property
and crops. In second is `r topten.totaldmg[2,1]` with `r topten.totaldmg[2,2]`
million dollars of damage each year and `r topten.totaldmg[2,4]` million dollars
of damage each event. In contrast to the events which caused the most harm with 
respect to public health, these events cause the most economic damage with very 
few events per year and as a result have a high economic cost per event. 